\begin{thebibliography}{9}
\bibitem{csp}
Communicating Sequential Processes
Available from: \url{http://www.usingcsp.com/cspbook.pdf}.
\bibitem{pat}
Process Analysis Toolkit (PAT) 3.5 User Manual
Available from: \url{https://pat.comp.nus.edu.sg/wp-source/resources/OnlineHelp/pdf/Help.pdf}.
\bibitem{blast}
The software model checker BLAST
Available from: \url{https://www.sosy-lab.org/research/pub/2007-STTT.The_Software_Model_Checker_BLAST.pdf}.
\bibitem{prism}
PRISM Model Checker
Available from: \url{https://www.prismmodelchecker.org/}.
\bibitem{tla}
Lamport L. The temporal logic of actions. ACM Transactions on Programming Languages and Systems (TOPLAS). 1994 May 1;16(3):872-923.
Available from: \url{https://lamport.azurewebsites.net/pubs/lamport-actions.pdf}.
\bibitem{tlaplus}
Lamport L. Specifying concurrent systems with TLA+. Calculational System Design. 1999 Apr 23:183-247.Available from: \url{https://lamport.azurewebsites.net/tla/xmxx01-06-27.pdf}.
\bibitem{tlc}
Yu Y, Manolios P, Lamport L. Model checking TLA+ specifications. InAdvanced Research Working Conference on Correct Hardware Design and Verification Methods 1999 Sep 27 (pp. 54-66). Berlin, Heidelberg: Springer Berlin Heidelberg.Available from: \url{https://lamport.azurewebsites.net/pubs/yuanyu-model-checking.pdf}.
\bibitem{pluscal}
Lamport L. The PlusCal algorithm language. InInternational Colloquium on Theoretical Aspects of Computing 2009 Aug 16 (pp. 36-60). Berlin, Heidelberg: Springer Berlin Heidelberg.Available from: \url{https://lamport.azurewebsites.net/pubs/pluscal.pdf}.
\bibitem{spin}
The Model Checker SPIN
Available from: \url{https://spinroot.com/spin/Doc/ieee97.pdf}.
\bibitem{go}
Build simple, secure, scalable systems with Go
Available from: \url{https://go.dev/}.
\bibitem{elixir}
Elixir is a dynamic, functional language for building scalable and maintainable applications.
Available from: \url{https://elixir-lang.org/}.
\bibitem{beam}
A brief introduction to BEAM
Available from: \url{https://www.erlang.org/blog/a-brief-beam-primer/}.
\bibitem{erlang}
Practical functional programming for a parallel world
Available from: \url{https://www.erlang.org/}.
\bibitem{phoenix}
Phoenix Peace of mind from prototype to production
Available from: \url{https://phoenixframework.org/}.
\bibitem{discord}
Discord
Available from: \url{https://discord.com/}.
\bibitem{ft}
Financial Times
Available from: \url{https://www.ft.com/}.
\bibitem{shared_memory_verification}
Mediero Iturrioz J. Verification of Concurrent Programs in Dafny.
Available from: \url{https://addi.ehu.es/bitstream/handle/10810/23803/Report.pdf?isAllowed=y&sequence=2}.
\bibitem{dafny}
The Dafny Programming and Verification Language
Available from: \url{dafny.org}
\bibitem{defmacro}
Elixir, Macros, Our First Macro
Available from: \url{https://hexdocs.pm/elixir/macros.html#our-first-macro}.
\bibitem{z3}
De Moura L, Bjørner N. Z3: An efficient SMT solver. InInternational conference on Tools and Algorithms for the Construction and Analysis of Systems 2008 Mar 29 (pp. 337-340). Berlin, Heidelberg: Springer Berlin Heidelberg.
Available from: \url{https://link.springer.com/content/pdf/10.1007/978-3-540-78800-3_24.pdf}.
\bibitem{hoare_logic}
Hoare CA. An axiomatic basis for computer programming. Communications of the ACM. 1969 Oct 1;12(10):576-80.
Available from: \url{https://dl.acm.org/doi/10.1145/363235.363259}
\bibitem{lean}
Lean and its Mathematical library
Available from: \url{https://leanprover-community.github.io/}.
\bibitem{dialyzer}
dialyzer
Available from: \url{https://www.erlang.org/doc/man/dialyzer.html}.
\bibitem{dafny_paper}
Leino KR. Dafny: An automatic program verifier for functional correctness. InInternational conference on logic for programming artificial intelligence and reasoning 2010 Apr 25 (pp. 348-370). Berlin, Heidelberg: Springer Berlin Heidelberg.
Available from: \url{https://link.springer.com/chapter/10.1007/978-3-642-17511-4_20}.
\bibitem{dafny_tutorial}
Nipkow T. Getting started with Dafny: A guide. Software Safety and Security: Tools for Analysis and Verification. 2012;33:152.
Available from: \url{https://dafny.org/dafny/OnlineTutorial/guide}.
\bibitem{boogie}
Barnett M, Chang BY, DeLine R, Jacobs B, Leino KR. Boogie: A modular reusable verifier for object-oriented programs. InFormal Methods for Components and Objects: 4th International Symposium, FMCO 2005, Amsterdam, The Netherlands, November 1-4, 2005, Revised Lectures 4 2006 (pp. 364-387). Springer Berlin Heidelberg.
Available from: \url{https://link.springer.com/chapter/10.1007/11804192_17}.
\bibitem{csp_paper}
Hoare CA. Communicating sequential processes. Englewood Cliffs: Prentice-hall; 1985 Jan.
\bibitem{distributed_algorithms_na_lynch}
Lynch NA. Distributed algorithms. Elsevier; 1996 Apr 16.
Available from: \url{https://lib.fbtuit.uz/assets/files/5.-NancyA.Lynch.DistributedAlgorithms.pdf}.
\bibitem{model_checking}
Clarke EM. Model checking. InFoundations of Software Technology and Theoretical Computer Science: 17th Conference Kharagpur, India, December 18-20, 1997 Proceedings 17 1997 (pp. 54-56). Springer Berlin Heidelberg.
\end{thebibliography}